---
title: "Final Project"
author: "Salcido"
date: "12/4/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.height=7, fig.width=12, fig.align="center")
```

## Exploratory Data Analysis on Attrition 
## DDS Final Project

We've obtained data on over 800 employees in a fortune 100 company and analyzed different factors that lead to attrition.

```{r}
## Load packages
# install.packages("ggplot2")
# install.packages("GGally")
# install.packages("dplyr")
# install.packages("tidyverse")
# install.packages("e1071")
# install.packages("class")

#Load Libraries
library(ggplot2)
library(GGally)
library(dplyr)
library(tidyverse)
library(e1071) 
library(class)

#Load data
CaseStudy <- read_csv(file.choose())
```

## Current Attrition

The first step in our exploratory data analysis is to measure and visualize current attrition in the company. We currently have a 19% Attrition rate in the company.

```{r}
EDA= CaseStudy
#Irrelevant Columns
drop= c("EmployeeCount", "EmployeeNumber", "Over18", "StandardHours", "Attrition.1")
EDA= EDA[ , !(names(EDA) %in% drop)]

#Graph Attrition
ggplot(EDA, aes(x="", y=Attrition, fill=Attrition)) +
  geom_bar(stat="identity", width=1) +
  ggtitle("Current Attrition Percent")+
  coord_polar("y")

#Attrition by Numbers
AttritionYes=nrow(EDA[EDA$Attrition == "Yes",])
AttritionNo=nrow(EDA[EDA$Attrition == "No",])
AttritionPercent= nrow(EDA[EDA$Attrition == "Yes",])/nrow(EDA[EDA$Attrition == "No",])
AttritionYes
AttritionNo
AttritionPercent
```
## Categorical Variables

For this step, the dataset has been broken down in two groups: Yes Attrition and No Attrition. 

We're looking for differences (in term of percent) between the two groups. If a variable has no impact in attrition it should have a similar percentage rate in both groups. If a substantial difference is seen between the groups, the variable may have an impact in attrition.

## Travel
No substantial difference is seen in travel.
The biggest difference on this category is seen in those who travel frequently. They have a larger percentage in the Attririon group than in the No Attrition group.

```{r}
#Travel
ggplot(EDA,mapping=aes(x=Attrition,fill=BusinessTravel))+
  geom_bar(position="fill")+
  ggtitle("Attrition by Travel")+
  ylab("Percentage")+
  scale_y_continuous(labels = scales::percent_format())
```

##Department

When we looked at differences between Departments we noticed that the Sales department has a substantial difference between the two Groups. They tend to have higher attrition rate. 

Research and Development has the lowest attrition.
```{r}
ggplot(EDA,mapping=aes(x=Attrition,fill=Department))+
  geom_bar(position="fill")+
  ggtitle("Attrition by Department")+
  ylab("Percentage")+
  scale_y_continuous(labels = scales::percent_format())
```

##EducationField
On the education field, those with a technical degree tend to have a higher attrion rate than the rest. 
```{r}
ggplot(EDA,mapping=aes(x=Attrition,fill=EducationField))+
  geom_bar(position="fill")+
  ggtitle("Attrition by EducationField")+
  ylab("Percentage")+
  scale_y_continuous(labels = scales::percent_format())
```

##Gender

After visualizing gender we can discard it as a factor of attrition.

```{r}
ggplot(EDA,mapping=aes(x=Attrition,fill=Gender))+
  geom_bar(position="fill")+
  ggtitle("Attrition by Gender")+
  ylab("Percentage")+
  scale_y_continuous(labels = scales::percent_format())
```

#JobRole
Job Role is an interesting variable since it has multiple categories. 
The job role that we will focus on is Sales Representative since it has a substantial difference between the attrition and no attrition group. 

Sales Representatives tend to be more prone to attritions than any other job role in the company.

```{r}
ggplot(EDA,mapping=aes(x=Attrition,fill=JobRole))+
  geom_bar(position="fill")+
  ggtitle("Attrition by JobRole")+
  ylab("Percentage")+
  scale_y_continuous(labels = scales::percent_format())
```

##MaritalStatus
Marital Status can be an important factor in attrition. Those employees who are single make up a substantially bigger portion in the attrition group than in the no attrition group.

On the contrary Divorced employees tend to be less likely to voluntarily resign. 
```{r}
ggplot(EDA,mapping=aes(x=Attrition,fill=MaritalStatus))+
  geom_bar(position="fill")+
  ggtitle("Attrition by MaritalStatus")+
  ylab("Percentage")+
  scale_y_continuous(labels = scales::percent_format())
```

##Overtime

Over Time is one of the variables with the highest difference among the attrition and no attrition groups.

Out of all the employees in the attrition group, 57% worked over time, while only 24% of employees that work over time make up the no attrition group. This is a group that we need to keep an eye on. 
```{r}
ggplot(EDA,mapping=aes(x=Attrition,fill=OverTime))+
  geom_bar(position="fill")+
  ggtitle("Attrition by OverTime")+
  ylab("Percentage")+
  scale_y_continuous(labels = scales::percent_format())
nrow(EDA[EDA$Attrition == "Yes" & EDA$OverTime == "Yes",])/nrow(EDA[EDA$Attrition == "Yes",])
nrow(EDA[EDA$Attrition == "No" & EDA$OverTime == "Yes",])/nrow(EDA[EDA$Attrition == "No",])
```
## Ordinal Variables Transformation
## Make Attrition Numeric
```{r}
NEDA = EDA[,c(3, 1:32)]
NEDA$Attrition=gsub('No', 0 , NEDA$Attrition)
NEDA$Attrition=gsub('Yes', 1 , NEDA$Attrition)
NEDA$Attrition= as.numeric(NEDA$Attrition) #Numeric
```

##Ordinal Variables Trends
We see an inverse realionship on most ordinal variables at the lower levels.

#Environment Satisfaction Visualized
```{r}
NEDA %>% ggplot(aes(EnvironmentSatisfaction, Attrition)) + geom_smooth()+
  ggtitle("Attrition and EnvironmentSatisfaction")
```
## Job Involvement
An inverse realtionship between Job involvement and attrition can be seen. 
```{r}
NEDA %>% ggplot(aes(JobInvolvement, Attrition)) + geom_smooth()+
  ggtitle("Attrition and JobInvolvement")
```
## JobLevel
Indirect relationship at the lower levels.
```{r}
NEDA %>% ggplot(aes(JobLevel, Attrition)) + geom_smooth()+
  ggtitle("Attrition and JobLevel")
```

## JobSatisfaction
There's an inverse relationship between Job Satisfaction and attrition.
```{r}
NEDA %>% ggplot(aes(JobSatisfaction, Attrition)) + geom_smooth()+
  ggtitle("Attrition and JobSatisfaction")
```
## Stock Option Level
 We can see that the inverse relationship between stock option level and attrition is only seen at the lower levels.
```{r}
NEDA %>% ggplot(aes(StockOptionLevel, Attrition)) + geom_smooth()+
  ggtitle("Attrition and Stock Option Level")
```
## WorkLifeBalance
Work life balance follows the same trend as other ordinal variables. The change is mostly present at the lower levels. 
```{r}
NEDA %>% ggplot(aes(WorkLifeBalance, Attrition)) + geom_smooth()+
  ggtitle("Attrition and WorkLifeBalance")
```

##Continuous Variables
 
 We continue visualizing our dataset to better understand the variables present.
 On this case we will use box plots to see any differences between the two groups in question.

##Age

This graph shows that the attrition group has a lower mean and median age than the no attrition group. 

```{r}
ggplot(EDA, aes(x=Attrition, y=Age, fill=Attrition)) + 
  geom_boxplot(notch=TRUE) +
  stat_summary(fun=mean, geom="point", shape=18, size=10, color="black", fill="red")+
  ggtitle("Attrition by Age") 
```

#DistanceFromHome

Distance from home shows that those who quit tend to leave farther away from their workplace. 

```{r}
ggplot(EDA, aes(x=Attrition, y=DistanceFromHome, fill=Attrition)) + 
  geom_boxplot(notch=TRUE) +
  stat_summary(fun=mean, geom="point", shape=18, size=10, color="black", fill="red")+
  ggtitle("Attrition by DistanceFromHome") 
```

##MonthlyIncome

Monthly Income shows that the attrition group tend to have a lower monthly income than the no attrition group. 

```{r}
ggplot(EDA, aes(x=Attrition, y=MonthlyIncome, fill=Attrition)) + 
  geom_boxplot(notch=TRUE) +
  stat_summary(fun=mean, geom="point", shape=18, size=10, color="black", fill="red")+
  ggtitle("Attrition by MonthlyIncome") 
```

##YearsInCurrentRole

It can be seen that the employees in the attrition group tend to have less time on their role before they leave than those who don't.
```{r}
ggplot(EDA, aes(x=Attrition, y=YearsInCurrentRole, fill=Attrition)) + 
  geom_boxplot(notch=TRUE) +
  stat_summary(fun=mean, geom="point", shape=18, size=10, color="black", fill="red")+
  ggtitle("Attrition by YearsInCurrentRole") 
```

##Predict Attrition

After visualizing our attrition factors and having a better understanding of our dataset we will attempt to predictr attrition by utilizing a couple of different predictive models. 

## Tidying up our Data for Prediction

In order to properly use our different kinds of variables to predict attrition we need to clean up and transform our data. 
```{r}
KNNEDA = EDA

#Drop Non Relevant Variables
drop= c("ID", "DailyRate", "Education", "Gender", "HourlyRate")
KNNEDA= KNNEDA[ , !(names(KNNEDA) %in% drop)]
KNNEDA$Attrition=factor(KNNEDA$Attrition)

#TransformCategorical Variables
KNNEDA$Non_Travel= ifelse(KNNEDA$BusinessTravel == "Non-Travel",1,0)
KNNEDA$Travel_Frequently= ifelse(KNNEDA$BusinessTravel == "Travel_Frequently",1,0) 
KNNEDA$Travel_Rarely= ifelse(KNNEDA$BusinessTravel == "Travel_Rarely",1,0) 

KNNEDA$HRDepartment= ifelse(KNNEDA$Department == "Human Resources",1,0)
KNNEDA$RDDepartment= ifelse(KNNEDA$Department == "Research & Development",1,0)
KNNEDA$SalesDepartment= ifelse(KNNEDA$Department == "Sales",1,0)

KNNEDA$HumanResourcesDegree= ifelse(KNNEDA$EducationField == "Human Resources",1,0)
KNNEDA$LifeSciencesDegree= ifelse(KNNEDA$EducationField == "Life Sciences",1,0)
KNNEDA$MarketingDegree= ifelse(KNNEDA$EducationField == "Marketing",1,0)
KNNEDA$MedicalDegree= ifelse(KNNEDA$EducationField == "Medical",1,0)
KNNEDA$OtherDegree= ifelse(KNNEDA$EducationField == "Other",1,0)
KNNEDA$TechnicalDegree= ifelse(KNNEDA$EducationField == "Technical Degree",1,0)

KNNEDA$HealthcareRep= ifelse(KNNEDA$JobRole == "Healthcare Representative",1,0)
KNNEDA$HumanResources= ifelse(KNNEDA$JobRole == "Human Resources",1,0)
KNNEDA$LaboratoryTechnician= ifelse(KNNEDA$JobRole == "Laboratory Technician",1,0)
KNNEDA$Manager= ifelse(KNNEDA$JobRole == "Manager",1,0)
KNNEDA$ManufacturingDirector= ifelse(KNNEDA$JobRole == "Manufacturing Director",1,0)
KNNEDA$ResearchDirector= ifelse(KNNEDA$JobRole == "Research Director",1,0)
KNNEDA$ResearchScientist= ifelse(KNNEDA$JobRole == "Research Scientist",1,0)
KNNEDA$SalesExec= ifelse(KNNEDA$JobRole == "Sales Executive",1,0)
KNNEDA$SalesRep= ifelse(KNNEDA$JobRole == "Sales Representative",1,0)

KNNEDA$Divorced= ifelse(KNNEDA$MaritalStatus == "Divorced",1,0)
KNNEDA$Married= ifelse(KNNEDA$MaritalStatus == "Married",1,0)
KNNEDA$Single= ifelse(KNNEDA$MaritalStatus == "Single",1,0)

KNNEDA$OverTime= ifelse(KNNEDA$OverTime == "Yes",1,0)

#Drop Transformed Variables
drop= c("BusinessTravel", "Department", "EducationField", "Gender", "JobRole", "MaritalStatus")
KNNEDA= KNNEDA[ , !(names(KNNEDA) %in% drop)]
KNNEDA= KNNEDA[,c(2, 1:46)]
KNNEDA= KNNEDA[ , !(names(KNNEDA)=="Attrition.1")]
```

## Top 3 Factors of Attrition

Once our data is ready we can determine the top 3 factors for attrition.
By running a generalized linear model and looking at factor's p value, we see that the top 3 Factors are: OverTime, JobInvolvement, JobSatisfaction

```{r}
model=glm(Attrition~.,data = KNNEDA, family= "binomial")
summary(model)
```
## KNN
The first model used to predict attrition is KNN. 
KNN gave us an 80% Accuracy, High Sensitivity, but low specificity.

```{r}
KNNEDA1= KNNEDA[, c(1,3,4,6, 11, 15,18, 23, 28, 34, 39, 43,46)]
set.seed(1)
confusionMatrix(table(knn.cv(KNNEDA1[,2:12],KNNEDA1$Attrition, k = 1), KNNEDA1$Attrition))
```


## Attrition-Naive Bayes
Our last model Naive Bayes gave us the best results. 
We achieved:
 86% Accuracy
 91% Sensitivity
 58% Specificity
```{r}
set.seed(1)
trainIndices = sample(1:dim(KNNEDA)[1],round(0.7 * dim(KNNEDA)[1]))
train = KNNEDA[trainIndices,]
test = KNNEDA[-trainIndices,]

model_train = naiveBayes(Attrition~.,data = train)
predict(model_train, train[,c(4,5,6,8,10,11,15,18,19,24,43,46)])

model_test = naiveBayes(Attrition~.,data = test)
classifications=predict(model_test, test[,c(4,5,6,8,10,11,15,18,19,24,43,46)])
CM_NaiveBayes=confusionMatrix(table(classifications,test$Attrition))
CM_NaiveBayes
```

## Interesting Finding and Final Recommendations
In our analysis we found it interesting that when it comes ordinal variables, the lowest level is the one that seems to have the most difference between those who stay and those who don't stay.

In general the Attrition group tends to have more employees who rank on the bottom level of the lowest level of the variable. We don't see that same difference on the rest of the levels.

Our recommendation is to focus on those employees that have no stock level, or have the lowest rank in Environment Satisfaction, Job Involvement, Job Level, Job Satisfaction, and work life balance.

If we can move those employees just one level higher we can have an impact on attrition. 

Below are some graphs visualizing those differences.


```{r}
ggplot(EDA,mapping=aes(x=Attrition,fill=as.character(EnvironmentSatisfaction)))+
  geom_bar(position="fill")+
  ggtitle("Attrition by EnvironmentSatisfaction")+
  ylab("Percentage")+
  scale_y_continuous(labels = scales::percent_format())
```
```{r}
ggplot(EDA,mapping=aes(x=Attrition,fill=as.character(JobInvolvement)))+
  geom_bar(position="fill")+
  ggtitle("Attrition by JobInvolvement")+
  ylab("Percentage")+
  scale_y_continuous(labels = scales::percent_format())
```
```{r}
ggplot(EDA,mapping=aes(x=Attrition,fill=as.character(JobLevel)))+
  geom_bar(position="fill")+
  ggtitle("Attrition by JobLevel")+
  ylab("Percentage")+
  scale_y_continuous(labels = scales::percent_format())
```
```{r}
ggplot(EDA,mapping=aes(x=Attrition,fill=as.character(JobSatisfaction)))+
  geom_bar(position="fill")+
  ggtitle("Attrition by JobSatisfaction")+
  ylab("Percentage")+
  scale_y_continuous(labels = scales::percent_format())
```
```{r}
ggplot(EDA,mapping=aes(x=Attrition,fill=as.character(StockOptionLevel)))+
  geom_bar(position="fill")+
  ggtitle("Attrition by StockOptionLevel")+
  ylab("Percentage")+
  scale_y_continuous(labels = scales::percent_format())
```
```{r}
ggplot(EDA,mapping=aes(x=Attrition,fill=as.character(WorkLifeBalance)))+
  geom_bar(position="fill")+
  ggtitle("Attrition by WorkLifeBalance")+
  ylab("Percentage")+
  scale_y_continuous(labels = scales::percent_format())
```
